Comparative analysis of PMV Models accuracy implemented in the ISO 7730:2005 and ASHRAE 55:2023

### 2025-01-28 Manuscript/src/sections/introduction.tex
review: addressed all reviewer's comments (with Ed's comments)

\todo[inline]{Ed's comment:
The PMV model consists of a heat balance model and an empirical equation that predicts thermal sensation (PMV) from heat balance.  (need to describe this further, especially the simplified way that sweating is treated by the heat balance model).
In 2010 ASHRAE Std 55 modified the PMV heat balance calculation to make it more capable of predicting the cooling caused by elevated air movement under warm conditions.
It uses the Gagge two-node model to compute the convective and evaporative heat loss from the body. This model explicitly models physiological cooling in warm conditions where evaporation of sweat becomes a significant source of heat loss. The model output can be expressed as Standard Effective Temperature (SET) (provide citations to Gagge/Foblets). SET is implemented in Std 55 as an index to quantify the equivalent temperature reduction caused by air movement cooling. The reduced temperature is entered into the PMV heat balance model in a still air condition, from which the PMV index is calculated. This approach is intended to minimize errors introduced by the Fanger heat balance model’s reduced evaporative capacity on the predicted PMV result.
}
\todo[inline, color=blue!40]{
    FT. Similar to my previous comment. Firstly, I would like to mention here that the PMV does not at all try to estimate sweating. This is a problem, but neither the PMV nor the PMV CE do that.
    The Std 55 did not modify the PMV heat balance equation, instead it simply and only reduced the air and mean radiant temperatures by the cooling effect, this is not equivalent to doing an accurate heat balance calculation.
    The PMV CE consequently does not: 1) use the Gagge two-node model to compute the convective and evaporative heat loss from the body and 2) model explicitly models physiological cooling in warm conditions where evaporation of sweat becomes a significant source of heat loss.
    In addition, another comment about the PMV CE is that the cooling effect is not only applied in warm conditions but is applied in any conditions.
    SET is also a rather confusing variable since it is calculated not for a reference environment with standardised clo and met. But rather it is the temperature of a hypothetical isothermal environment at 50\% rh, <0.1 m/s (20 fpm) average air speed Va, and in which the total heat loss from the skin of an imaginary occupant wearing clothing, standardized for the activity concerned, is the same as that from a person in the actual environment with actual clothing and activity level.
    Finally, as you mentioned, \textit{The reduced temperature is entered into the PMV heat balance model in a still air condition, from which the PMV index is calculated.} this however does not ensure that the person is not feeling hot in this condition or sweating, consequently does not solve the issues that the original PMV has.
}

\subsection{\ac{pmv-ce} model}\label{subsec:pmv-ce-limitations}
The \gls{55} standard uses a modified version of the original model, the \ac{pmv-ce}~\cite{ashrae552023}.
The best source to see how \ac{pmv-ce} works is the standard itself, but the standard does not explain why some changes to the \ac{pmv} inputs were implemented, nor for example, why the \ac{ce} is subtracted from the \ac{tdb} and \ac{tr}.
A partial justification of the model is mainly described in \mycite{arens_moving_2009} and secondarily in \mycite{yang_cooling_2015}.
However, we could not find a peer-reviewed scientific publication that quantified the accuracy improvements of the model as implemented in the \gls{55} over the original \ac{pmv} model.
\todo[inline]{true enough, but in the standard, the objective is to have a rational physiology model.
The Gagge models are as well or better documented than the Fanger heat balance model, and have been available to Standards users in the ASHRAE Handbook of Fundamentals.}
\todo[inline, color=blue!40]{FT. The PMV CE and the Gagge model are very different, hence, publications about the latter do not justify the implementation proposed in the Std 55}

@@

To illustrate the extent of the differences between the outputs of the two models, only as a function of \ac{vr}, we calculated the comfort regions ($\mid$\ac{pmv}$\mid \leq 0.5$) assuming the same \ac{met}~=~\qty{1.2}{met} and \ac{clor}~=~\qty{0.5}{clo}.
We plotted the results in Figure~\ref{fig:comfort_regios_pmv_pmvce}.
\todo{It would be great to add 0.6m/s to this figure. That would be the objective for ceiling fan designs. Many lab studies have measured results in that region.}
\todo[color=blue!40]{FT. I can include a Figure for 0.6m/s but I did not since based on the ASHRAE DB 2 we hardly reach that value in buildings.}
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{figures/pmv_comfort_regions}
    \caption{Comfort regions ($|$\ac{pmv}$|$~$\leq$~\num{0.5}) calculated using \ac{pmv} and \ac{pmv-ce} models for two values of \ac{vr}.
    We set \ac{met}~=~\qty{1.2}{met}, \ac{clor}~=~\qty{0.5}{clo}.
    \label{fig:comfort_regios_pmv_pmvce}}
\end{figure}
The results show that for \ac{vr}~=~\qty{0.4}{\m\per\s} and \ac{rh}~=~\qty{50}{\percent} the comfort regions estimated using the \ac{pmv-ce} is \qty{0.7}{\celsius} wider than the one calculated using the \ac{pmv} and shifted towards warmer temperatures.
For \ac{vr}~=~\qty{0.2}{\m\per\s}, the same difference is \qty{0.4}{\celsius}.

@@

    \item Although the \ac{ce} claims to better account for convective heat losses from the skin to the environment due to the air movement, the \ac{pmv-ce} model uses the same underlying equations as the \ac{pmv} model, hence it has the same limitations.
    For example, let's assume that a person is exposed to the following environmental conditions \ac{tdb}~=~\ac{tr}~=~\qty{29}{\celsius}, \ac{rh}~=~\qty{80}{\percent}, \ac{vr}~=~\qty{0.15}{\m\per\s}, they have a \ac{met} of \qty{1}{met} and a \ac{clor} of \qty{0.5}{clo}.
    According to Gagge's two-node model, which is used to calculate the \ac{set}, this hypothetical person would have \qty{30}{\percent} of their skin covered with sweat.
    Based on the results from \mycite{huang_applicability_2014}, the \ac{pmv} model, in this scenario, cannot accurately predict the \ac{tsv} since the person is sweating.
    However, the only difference between the \ac{pmv} and \ac{pmv-ce} in this scenario, is that the \ac{pmv-ce} accounts for the cooling effect due to the air movement (delta between \qty{0.15}{\m\per\s} and \qty{0.1}{\m\per\s}) which is equal to \qty{.17}{\celsius}.

@@

    You are citing skin wettedness values that have not been introduced in this paper (and are not part of any standard). Where did they come from? You allude to results from Huang 2014 that the reader is supposed to know about. The skin-wettedness discussion is irrelevant unless you go into a lot more physiological detail here.}

@@

    \item The \ac{set} temperature which is used to calculate the \ac{ce}, is the hypothetical isothermal environment at a \ac{rh} of \qty{50}{\percent}, \ac{vr}~$\leq$~\qty{0.1}{\m\per\s}, and \ac{tr}~=~\ac{tdb} in which the total heat loss from the skin of an imaginary occupant wearing clothing, standardized for the activity concerned, is the same as that from a person in the actual environment with actual clothing and activity level~\cite{ashrae552023}.
    However, while \ac{vr}, used to calculate the \ac{pmv-ce}, is set to \qty{0.1}{\m\per\s} the value of \ac{rh} and \ac{clo} are not adjusted to compensate for the fact that \ac{set} fixed the \ac{rh}~=~\qty{50}{\percent} and standardized the clothing to the activity.
    \todo[inline]{I think this argument is incorrect. SET is only used as a consistent method of obtaining a meaningful delta operative temperature. It does not change the environment. The environment is kept at the original RH.}

@@

@@

>>>>

\todo[inline]{Ed's comment:
The PMV model consists of a heat balance model and an empirical equation that predicts thermal sensation (PMV) from heat balance.  (need to describe this further, especially the simplified way that sweating is treated by the heat balance model).
In 2010 ASHRAE Std 55 modified the PMV heat balance calculation to make it more capable of predicting the cooling caused by elevated air movement under warm conditions.
It uses the Gagge two-node model to compute the convective and evaporative heat loss from the body. This model explicitly models physiological cooling in warm conditions where evaporation of sweat becomes a significant source of heat loss. The model output can be expressed as Standard Effective Temperature (SET) (provide citations to Gagge/Foblets). SET is implemented in Std 55 as an index to quantify the equivalent temperature reduction caused by air movement cooling. The reduced temperature is entered into the PMV heat balance model in a still air condition, from which the PMV index is calculated. This approach is intended to minimize errors introduced by the Fanger heat balance model’s reduced evaporative capacity on the predicted PMV result.
}
\todo[inline, color=blue!40]{
    FT. Firstly, it is important to note that the PMV does not aim to estimate sweating. 
    While this is a limitation, neither the PMV nor the PMV CE models include this capability.  
    The Std 55 did not modify the PMV heat balance equation but instead reduced the air and mean radiant temperatures by the cooling effect, assuming this adequately accounts for sweating. 
    However, this approach does not equate to an accurate heat balance calculation where sweat rate is explicitly modeled, as seen in the Gagge two-node model.  
    Consequently, the PMV CE does not: 1) use the Gagge two-node model to compute convective and evaporative heat loss from the body, nor 2) explicitly model physiological cooling in warm conditions where sweat evaporation becomes a key source of heat loss.  
    Secondly, in the PMV CE model, the cooling effect is applied in all conditions, not just in warm ones.  
    Thirdly, the definition of SET can also be somewhat unclear. 
    It is not calculated for a reference environment with standardised clothing and metabolic rates. 
    Instead, it represents the temperature of a hypothetical isothermal environment at 50\% RH, $<$0.1 m/s (20 fpm) average air velocity, where the total heat loss from the skin of an imaginary occupant (wearing clothing standardised for the activity) equals the heat loss from a person in the actual environment with their actual clothing and activity level.  
    Finally, as you mentioned, \textit{The reduced temperature is entered into the PMV heat balance model in a still air condition, from which the PMV index is calculated.} However, this does not guarantee that the person will not feel hot or sweat in this condition and, therefore, does not address the inherent issues of the original PMV model.
}

\subsection{\ac{pmv-ce} model}\label{subsec:pmv-ce-limitations}
The \gls{55} standard uses a modified version of the original model, the \ac{pmv-ce}~\cite{ashrae552023}.
The best source to see how \ac{pmv-ce} works is the standard itself, but the standard does not explain why some changes to the \ac{pmv} inputs were implemented, nor for example, why the \ac{ce} is subtracted from the \ac{tdb} and \ac{tr}.
A partial justification of the model is mainly described in \mycite{arens_moving_2009} and secondarily in \mycite{yang_cooling_2015}.
However, we could not find a peer-reviewed scientific publication that quantified the accuracy improvements of the model as implemented in the \gls{55} over the original \ac{pmv} model.
\todo[inline]{EA. true enough, but in the standard, the objective is to have a rational physiology model.
The Gagge models are as well or better documented than the Fanger heat balance model, and have been available to Standards users in the ASHRAE Handbook of Fundamentals.}
\todo[inline, color=blue!40]{FT. The PMV CE and the Gagge model are different, hence, publications about the latter do not sufficiently describe the rationale of the new model, moreover, the accuracy of the PMV CE was not tested}

@@

To illustrate the extent of the differences between the outputs of the two models, only as a function of \ac{vr}, we calculated the comfort regions ($\mid$\ac{pmv}$\mid \leq 0.5$) assuming the same \ac{met}~=~\qty{1.2}{met} and \ac{clor}~=~\qty{0.5}{clo}.
We plotted the results in Figure~\ref{fig:comfort_regios_pmv_pmvce}.
\todo{It would be great to add 0.6m/s to this figure. That would be the objective for ceiling fan designs. Many lab studies have measured results in that region.}
\todo[color=blue!40]{FT. In the manuscript, we decided not to include a Figure for 0.6m/s since in the ASHRAE DB 2 we hardly reach that value.
I am happy to generate a figure for you if need be}
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{figures/pmv_comfort_regions}
    \caption{Comfort regions ($|$\ac{pmv}$|$~$\leq$~\num{0.5}) calculated using \ac{pmv} and \ac{pmv-ce} models for two values of \ac{vr}.
    We set \ac{met}~=~\qty{1.2}{met}, \ac{clor}~=~\qty{0.5}{clo}.
    \label{fig:comfort_regios_pmv_pmvce}}
\end{figure}
The results show that for \ac{vr}~=~\qty{0.4}{\m\per\s} and \ac{rh}~=~\qty{50}{\percent} the comfort regions estimated using the \ac{pmv-ce} is \qty{0.7}{\celsius} wider than the one calculated using the \ac{pmv} and shifted towards warmer temperatures.
For \ac{vr}~=~\qty{0.2}{\m\per\s}, the same difference is \qty{0.4}{\celsius}.

@@

    \item Although the \ac{ce} claims to better account for the heat losses from the skin to the environment due to the air movement, the \ac{pmv-ce} model uses the same \ac{pmv} model equations, hence it has the same limitations.
    This can be shown with a thought-example.
    Let's assume that a person is exposed to the following environmental conditions \ac{tdb}~=~\ac{tr}~=~\qty{29}{\celsius}, \ac{rh}~=~\qty{80}{\percent}, \ac{vr}~=~\qty{0.25}{\m\per\s}, they have a \ac{met} of \qty{1}{met} and a \ac{clor} of \qty{0.5}{clo}.
    In these conditions, a person will be sweating; in fact, according to Gagge's two-node model, which is used to calculate the \ac{set}, a person would have approximately \qty{30}{\percent} of their skin covered with sweat.
    The \ac{pmv} model, in this scenario, cannot accurately predict the \ac{tsv} since the person is sweating \mycite{huang_applicability_2014}.
    However, the only difference between the \ac{pmv} and \ac{pmv-ce} in this thought example is that the \ac{pmv-ce} accounts for the cooling effect due to the air movement (delta between \qty{0.25}{\m\per\s} and \qty{0.1}{\m\per\s}) which is equal to \qty{1.1}{\celsius}.

@@

    You are citing skin wettedness values that have not been introduced in this paper (and are not part of any standard). 
    Where did they come from? You allude to results from Huang 2014 that the reader is supposed to know about. 
    The skin-wettedness discussion is irrelevant unless you go into a lot more physiological detail here.}
    \todo[inline, color=blue!40]{FT. We developed this thought-example to demonstrate that while the PMVCE seeks to improve the modeling of sweat rates, it does not alter the underlying PMV equations. 
    As a result, there are situations where the PMVCE may still fail to provide accurate predictions.
    We have modified the example and changed the air speed from 0.15 to 0.25 m/s}

@@

@@

    In some contexts, like in yearly building performance simulations, this could make a difference. 

@@

    \todo[inline, color=blue!40]{FT. I agree that this may or may not be significant, depending on the context and the computation speed could be improved.}

### 2025-01-28 Manuscript/src/sections/results.tex
review: addressed all reviewer's comments (with Ed's comments)

\subsection{Sources of Error}\label{subsec:sources-of-error}
In this section, we try to identify some of the main reasons that could explain the difference in the results from the two \ac{pmv} models when compared against the \ac{tsv} available in the \ac{db2}.
Moreover, we tried to explain why we observed that the \ac{pmv-ce} model has a lower accuracy than the \ac{pmv} in predicting thermal sensation.
The issues are reported in order of importance of what we believe may affect the accuracy of the model.
\todo{Ed's comment - Note that the section below does not mention the two-node model, the CE implementation of PMV, or the physiological differences between the ISO and ASHRAE models that could explain the observed differences or ‘error’. }

\paragraph{Heat Balance Equation and Calculation of the PMV Value}
The \ac{pmv} model uses simplified heat balance equations to estimate the heat losses and gains from the human body to its surrounding environment.
This is a possible source of error since the model considers the human body to be a cylinder with constant width all uniformly covered with clothing where the heat is all generated in its core.
This is a significant simplification, as it overlooks several important factors, such as the uneven coverage of clothing across different parts of the body, the non-uniformity of metabolic heat generation, and the variations of mass-to-surface ratios.
It also ignores vasoconstriction and vasodilation.
Additionally, the \ac{pmv} model erroneously assumes that the human body is not capable of maintaining a stable core temperature.
A \ac{pmv} value higher than \num{.5} or lower than \num{-.5} indicate that the hypothetical cylinder representing the human body is either gaining or losing heat, and consequently it is getting warmer or colder, respectively.

@@

The \ac{pmv} model also assumes that the human body is constantly losing a certain amount of heat through sweating and this heat loss (W/m\textsuperscript{2}) only varies as a function of \ac{met} and it is calculated using the following equation $0.42\times($\ac{met}$\times58.15 - 58.15)$~\cite{Fanger1970}.
This amount of heat loss is included in the equation even if the cylinder has a negative heat gain, or in other words, the person is estimated to be feeling cold.

The \ac{pmv} model uses the overall heat losses or gains to calculate a \ac{pmv} value which should represent the average \ac{tsv} of a large group of occupants.
Heat losses or gains are only a proxy for the thermoregulation effort that the human body has to undergo to maintain a constant internal temperature.
Fanger created this equation based on a limited dataset collected in a laboratory, and this equation was never updated or revisited~\cite{Fanger1970}.
In the \ac{pmv} model the heat losses and gains are mapped using constant coefficients to a \ac{tsv}, disregarding the fact that the human body employs different strategies, i.e., vasodilation and vasoconstriction to cope with warm and cold conditions~\cite{romanovsky_thermoregulation_2018}.

>>>>

\subsection{Sources of Error}\label{subsec:sources-of-error}
In this section, we try to identify some of the main reasons that could explain the difference in the results from the two \ac{pmv} models when compared against the \ac{tsv} available in the \ac{db2}.
The issues are reported in order of importance of what we believe may affect the accuracy of the model. 
In this paragraph, we refer to the \ac{pmv} model, but the consideration can also be applied to the \ac{pmv-ce} model given its formulation.

\paragraph{Heat Balance Equations and Thermal Sensation Predictions}
The \ac{pmv} model uses simplified heat balance equations to estimate the heat losses and gains from the human body to its surrounding environment.
This is a possible source of error since the model considers the human body to be a cylinder with constant width, all uniformly covered with clothing, where the heat is all generated in its core.
This is a significant simplification, as it overlooks several important factors, such as the uneven coverage of clothing across different parts of the body, the non-uniformity of metabolic heat generation, and the variations of mass-to-surface ratios.
It also ignores vasoconstriction and vasodilation.
Additionally, the \ac{pmv} model erroneously assumes that the human body is not capable of maintaining a stable core temperature.
A \ac{pmv} value higher than \num{.5} or lower than \num{-.5} indicate that the hypothetical cylinder representing the human body is either gaining or losing heat, and consequently, it is getting warmer or colder, respectively.

@@

The \ac{pmv} model also assumes that the human body is constantly losing a certain amount of heat through sweating and this heat loss (W/m\textsuperscript{2}) only varies as a function of \ac{met} and it is calculated using the following equation $0.42\times($\ac{met}$\times58.15 - 58.15)$~\cite{Fanger1970}.
This amount of heat loss is included in the equation even if a person is in thermal neutrality (feeling comfortable) or has a negative heat gain (feeling cooler than neutral).

The \ac{pmv} model uses the overall heat losses or gains to calculate a \ac{pmv} value which should represent the average \ac{tsv} of a large group of occupants.
Heat losses or gains are only a proxy for the thermoregulation effort that the human body has to undergo to maintain a constant internal temperature.
Fanger created this equation based on a limited dataset collected in a laboratory, which was never updated or revisited~\cite{Fanger1970}.
In the \ac{pmv} model, the heat losses and gains are mapped using constant coefficients to a \ac{tsv}, disregarding the fact that the human body employs different strategies, i.e., vasodilation and vasoconstriction, to cope with warm and cold conditions~\cite{romanovsky_thermoregulation_2018}.
